---
title: "Hops Project_MB599"
author: "Agung Irawan"
date: "4/25/2022"
output: html_document
---

## Including Plots
```{r}
library(ggplot2)
library(vegan)
library(permute)
library(dplyr)
library(magrittr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(randomForest)
library(knitr)
library(caret)
library(glmnet)
```

#Load data
```{r}
psdata <- readRDS("~/irawana_mb599/phyloseq/mb599data.Rds")
View(as.matrix.data.frame(sample_data(psdata)))

```

# Factorize and set the levels on the metadata
```{r}
week_levels <- c("1", "2", "3", "4", "5")
sample_data(psdata)$Week = factor(week_levels)

ids <- sample_data(psdata)$Participant.ID
participant_levels <- c("101", "102", "103", "104", "105", "106", "107", 
                        "109", "110", "111", "112", "113", "114", "115", 
                        "116", "117", "118", "119", "120", "121", "122", 
                        "123", "124" ,"125", "126", "127", "128", "129", 
                        "130", "131")
sample_data(psdata)$Participant.ID <-  factor(ids, levels = participant_levels)

group_levels <- c("treatment", "control")
group <- get_variable(psdata, "Group")
sample_data(psdata)$Group = factor(group, levels = group_levels)

# Agglomerate to the genus level
ps_genera <- psdata %>% tax_glom(taxrank = "Genus")
# Remove taxa not seen more than 3 times in at least 20% of the samples
ps_counts <- ps_genera %>% filter_taxa(function(x) sum(x > 3) > (0.2*length(x)), TRUE)
# Convert from counts to relative abundance
ps_relab <- ps_counts %>% transform_sample_counts(function(x) x / sum(x))
# Filter out low abundance (>1e-5) taxa
ps <- ps_relab %>% filter_taxa(function(x) mean(x) > 1e-5, TRUE)

# Add Metabotypes for DXN and 8PN
sample_data(psdata)$metabo_dxn <- ifelse(sample_data(psdata)$DXN > 2292, "high", "low")
sample_data(psdata)$metabo_8pn <- ifelse(sample_data(psdata)$X8PN > 1910, "high", "low")
# Facotize DXN and 8PN metabotypes
metabotypes <- c("high", "low")
sample_data(psdata)$metabo_dxn <- factor(sample_data(psdata)$metabo_dxn, levels = metabotypes)
sample_data(psdata)$metabo_8pn <- factor(sample_data(psdata)$metabo_8pn, levels = metabotypes)
```

#PCoA
```{r, fig.height = 3}
ord <- ordinate(ps, method = "PCoA", distance = 'euclidean')
plot_ordination(ps, ord, 'samples', color = 'Group')
```

```{r}
par(mfrow=c(1,2))
ps@sam_data$Group_treatment = ps@sam_data$Group == "treatment"
ps@sam_data$Group_control = ps@sam_data$Group == "control"
plot_ordination(ps, ord, 'samples', color = 'Group_treatment')
plot_ordination(ps, ord, 'samples', color = 'Group_control')
```

#Random forest
```{r}
set.seed(1)
index_train <- createDataPartition(psdata@sam_data$Group, p = 0.5)[[1]]
train <- psdata@otu_table[index_train, ]
test <- psdata@otu_table[-index_train, ]

#spliting the phyloseq objects
pstrain <- phyloseq(otu_table(train, taxa_are_rows = FALSE), psdata@sam_data[index_train, ])
pstest <- phyloseq(otu_table(test, taxa_are_rows = FALSE), psdata@sam_data[-index_train, ])

#Train model
model_ridge <- glmnet(x = as.matrix(train), y = pstrain@sam_data$Group, family = 'binomial', alpha = 0)
model_ridge
```


#Find optimum mtry
```{r}
set.seed(1)
datatrain = data.frame(train)
datatrain$sample_group = pstrain@sam_data$Group
control <- trainControl(method='repeatedcv', 
                        number=3, 
                        repeats=3,
                        allowParallel = F)

tunegrid <- expand.grid(.mtry=c(1:20)) 
rf <- train(sample_group ~., 
            data= datatrain, 
            method='rf', 
            metric='Accuracy', 
            tuneGrid=tunegrid, 
            trControl=control)
print(rf)
```


#Testing model performance
```{r}
mtry_best = as.numeric(rf$bestTune)
model = randomForest(train, y = as.factor(pstrain@sam_data$Group), mtry = mtry_best)

#Performance on test set
preds = predict(model, test)
print(paste("Accuracy: ", sum(preds == as.factor(pstest@sam_data$Group)) / nsamples(pstest)))

#Visualize on test dataset
ord <- ordinate(pstest, method = "PCoA", distance = 'euclidean')
pstest@sam_data$rf_predictions = predict(model, pstest@otu_table)
plot_ordination(pstest, ord, 'samples', color = 'Group', shape = 'rf_predictions') + geom_point(size = 4) +
  ggtitle("Microbiome classification between groups based on random forest algorithm")
```

#obtain the important taxa based on treatment vs control group
```{r}
model = randomForest(pstest@otu_table, y = as.factor(pstest@sam_data$Group), mtry = mtry_best)
varImpPlot(model, type = 2)
```


#obtain the important taxa based on DXN metabolite concentration "High vs Low"
```{r}
model = randomForest(pstest@otu_table, y = as.factor(pstest@sam_data$metabo_dxn), mtry = mtry_best)
varImpPlot(model, type = 2)
```


##obtain the important taxa based on 8PN metabolite concentration "High vs Low"

```{r}
model = randomForest(pstest@otu_table, y = as.factor(pstest@sam_data$metabo_8pn), mtry = mtry_best)
varImpPlot(model, type = 2)
```


#Important variables
```{r}
imp_list <- list()
for(i in 1:20){
  model = randomForest(pstest@otu_table, y = as.factor(pstest@sam_data$Group), mtry = mtry_best)
  imp_list[i] <- varImp(model)
}

imp_df <- do.call(rbind.data.frame, imp_list)
colnames(imp_df) <- colnames(pstest)
colMeans(imp_df)
barplot(sort(colMeans(imp_df)), horiz = T, las = 1, xlab = "Mean variable importance")
```


